Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-17 article.tex
Signal filtering

In this paper, we describe the winning method\footnote{Code available at
\url{https://github.com/asutera/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics to infer the neural network. The rest of the
paper is structured as follows: Section \ref{sec:filter} describes the signal
processing methods applied on fluorescent calcium time series; Section
\ref{sec:inference} then presents the proposed approach and its theoretical
properties; Section \ref{sec:results} provides an empirical analysis and
comparison with other network inference methods, while Section
\ref{sec:conclusion} finally discusses our work and provides further research
directions.

@@

Given the image acquisition process, the fluorescent calcium time series
are smoothed using several linear and non linear filters in
order to reduce the imaging process effects. The overall filtering approach
is displayed at Figure \ref{fig:filtered-signal}.

\begin{figure}[bth]
\centering
\includegraphics[width=0.4\linewidth]{images/fig_filtering}
\caption{Filtering steps.}
\label{fig:filtered-signal}
\end{figure}

The raw signal (original signal of Figure \ref{fig:filtered-signal}) is very noisy
due to light scattering artifacts that ordinarily affects the quality of the
recording (\cite{lichtman2011big}). A low-pass filter is first applied to
filter high frequency noises with one of the following designs
\begin{align}
% symmetrical median filter
x^t_i &\leftarrow x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
x^t_i &\leftarrow 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i.
\label{eq:weighted-asymetric-median}.
\end{align}

A neuron spike lasts on average $1ms$ to $2ms$, while current optical
device have a time resolution of $20ms$. Considering the short delay for a
communication between neurons and the slow decay of fluorescence signal, we
filter low frequencies through a high pass filter
\begin{align}
% an asymmetrical discrete derivative filter
x^{t,d}_{i} &\leftarrow x^{t}_i - x_{t-1}^i. \label{eq:high-pass-filter}
\end{align}

Whenever the low and high pass filtered have been applied, the remaining signal
(H-P signal of Figure \ref{fig:filtered-signal}) is the discrete
derivative of the fluorescence calcium concentration. To further filter out
small signal variations which are due to noise, we apply a hard-threshold
filter
\begin{align}
x^{t,d}_i &\leftarrow x^{t,d}_i \mathbb{1}(x^{t,d}_i \geq \tau) \text{ with } \tau > 0.
\end{align}

After those 3 first filters, the fluorescence signal (Thresh. signal of
Figure \ref{fig:filtered-signal}) only contains clean spikes. However when a large
part of the network is firing, it's difficult to detect pairwise neuron
interactions. In order to reduce false interaction detection,
the signal is smoothed with a non-linear filter which penalizes
the filtered signal when a large part of the network is firing
\textcolor{red}{This equation doesn't correspond to the what we have done.}
\begin{align}
 x^{t,w}_i  &\leftarrow (x^{t,d}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t,d}_j}}.
\end{align}
Once this filtered has been applied (Weights signal of
Figure \ref{fig:filtered-signal}), peak intensities have
changed depending on the overall network activity, but not their numbers and
locations.

\textcolor{red}{PCA?} Finally since there is a high correlation between neurons,
we filter some more noise through a principal component analysis. From the $|V|$
eigen vectors, we kept $80\%$ of the eigen vectors which have the largest eigen
value. In the context of the challenge, this assumption was deduced from ground
truth of provided networks.

@@

If the random variables $X$ follows a joint Gaussian distribution
$p_\mathcal{X} \sim \mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and
covariance matrix $\Sigma$, we have the conditional independence
$X_i \perp X_j | X^{-i,j}$ if the partial correlation coefficient
\[
\rho_{X_i, X_j | X^{-i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}
\]
is equal to zero (\cite{koller2009probabilistic}). The partial correlation gives
a connectivity score
$\hat{A}: \hat{a}_{i,j} = \rho_{X_i, X_j | X^{-i,j}}, \forall i, j \in V$ between all
pairs of neuron $i$ and $j$. One could retrieve an estimated graph by thresholding
the matrix $\hat{A}$.
Similarly, we also have $X_i \perp X_j$ if the Pearson correlation
coefficient
\[
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}}
\]
is equal to zero.

@@

% Introduce a custom solution to make the Y matrix asymmetric
% Introduce directivity
\textcolor{red}{Est-ce qu'on ne le ferait pas passer dans l'appendice B? Ou du moins l'équation?}
Since partial correlation is a symmetric measure, the causal mechanism behind the
interaction ($A$ directly causes $B$, $B$ directly causes $A$ or both) can not
be directly inferred\footnote{Also note that two other causal mechanisms might be
implied while having a non-zero partial coefficient: (i) a pair of variables
is induced by a common (hidden) variable; (ii) $A$ (respectively $B$) is
conditionally correlated to a hidden variable affecting $B$ (resp. $A$)
\cite{de2004discovery}.}.
In order to retrieve causal information, we developed an
heuristics based on the following observation. The rise of fluorescence
indicates the activation of a neuron $i$. If a neuron $j$ have also
an increased of fluorescence after a slight delay, this could have been
caused by the neuron $i$. Thus, we count whenever
such events occurred between consecutive time steps
\[
\hat{Z}: z_{ij} = \sum_{t=1}^{T - 1}
    \mathbb{1}(x_j^{t+1} - x_i^t \in \left[f_1, f_2 \right]) -
    \mathbb{1}(x_i^{t+1} - x_j^t \in \left[f_1, f_2 \right]), \forall i, j \in V
\]
where $f_1$ and $f_2$ are parameters of the method.
Finally, the weighted average of correlation matrices is combined with $Z$ such
that $Z$ has a contribution of $0.3\%$ to the prediction of an edge.

@@

% Datasets
To assess our method, we have used four datasets provided by the organizers
of the Connectomics challenge : \textit{normal-1, normal-2, normal-3, normal-4}.
Those datasets are time series of size $T=179500$ of image calcium from simulated
neuronal networks (\cite{stetter2012model}) with $|V|=1000$ neurons. The networks
to inferred have around 15000 edges and are similar to those
used to rank challenge participants.

% Metrics : ROC AUC and AUPRC
In supervised learning terms, the network inference task can be viewed as a
binary classification task where one has to correctly classify the presence
or the absence of edges of the graph. We assess the accuracy of the
network inference method using the area under the ROC curve (AUROC)
and the area under the precision-recall curve (AUPRC)
(\cite{schrynemackers2013protocols}). Since neurons are not self-connected
($(i, i) \not \in E, \forall i \in V$) and some inference methods could
give maximal scores to the diagonal of $\hat{A}$, e.g. Pearson
correlation, we set those elements to the minimum of $\hat{A}$ prior
assessment.

@@

GTE is a method which characterizes a relationship between neurons by an
estimation of the generalized transfer entropy (\cite{stetter2012model}). The
goal of this method tries to see if the next time step value is better from
the previous values of this time series or from time series of other neurons.

@@

In previous study \cite{shipley2002cause}, it has been suggested to
consider correlation coefficients of all possible orders,
i.e. conditioned on every subsets of the set of $p-2$ other variables. This
would take into account multiple indirect paths between a given pair of
variables.

>>>>

In this paper, we describe the winning method\footnote{Code available at
\url{https://github.com/asutera/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The rest of the
paper is structured as follows: Section \ref{sec:filter} describes the signal
processing methods applied on fluorescent calcium time series; Section
\ref{sec:inference} then presents the proposed approach and its theoretical
properties; Section \ref{sec:results} provides an empirical analysis and
comparison with other network inference methods, while Section
\ref{sec:conclusion} finally discusses our work and provides further research
directions.

@@

Under the simplifying assumption that neurons are on-off units, characterized
by short periods of intense activities, or peaks, and longer periods of
inactivity, the first part of our algorithm consists in cleaning the raw
fluoresence data. More specifically, time series are processed using standard
signal processing filters in order to remove noise due light scattering
effects, account for fluorescence  low decay and reduce the importance of
global high activity in the network. The overall process is illustrated on
Figure~\ref{fig:filtered-signal}.

\begin{figure}
\centering
\includegraphics[width=0.4\linewidth]{images/fig_filtering}
\caption{Signal processing process.}
\label{fig:filtered-signal}
\end{figure}

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, due to
light scattering artifacts that ordinarily affects the quality of the
recording~\citep{lichtman2011big}, the raw fluoresence signal is very noisy.
Accordingly, the first step of our pipeline is to use a low-pass filter for
filtering out high frequency noise, using one of the following designs,
resulting in the second plot in the figure:
\begin{align}
% symmetrical median filter
x^t_i &\leftarrow x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
x^t_i &\leftarrow 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i.
\label{eq:weighted-asymetric-median}
\end{align}

Considering the short delay of communication between neurons and the slow decay
of fluorescence signal, the second step of our pipeline consists in identifying
neuron spikes -- characterized by high frequencies -- by filtering out low
frequencies through a high-pass filter, resulting in third plot of the figure:
\begin{align}
% an asymmetrical discrete derivative filter
x^{t,d}_{i} &\leftarrow x^{t}_i - x_{t-1}^i. \label{eq:high-pass-filter}
\end{align}

Once the low- and and high-pass filters have been applied, the remaining signal
corresponds to the discrete derivative of the fluorescence time series. To
further filter out small signal variations, mostly due to noise, the third step
consists in a  hard-threshold filter, yielding the fourth plot in the figure:
\begin{align}
x^{t,d}_i &\leftarrow x^{t,d}_i \mathbb{1}(x^{t,d}_i \geq \tau) \text{ with } \tau > 0.
\end{align}

At this step, the processed signal only contains clean spikes. However, when a
large of the network is firing, detecting pairwise neuron interactions is a
difficult task. \textcolor{red}{G: say that neurons may fire because of global
high activity, without any actual cause -- see generator.}
Accordingly, in order to reduce false interaction detection,
peak intensities at time $t$ are regularized by the global activity of all
neurons, thereby reducing the importance of peaks identified during global high activity periods:
\begin{align}
 x^{t,w}_i  &\leftarrow (x^{t,d}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t,d}_j}}.
\end{align}

\textcolor{red}{G: Let's drop that? or reformulate...}
\textcolor{red}{PCA?} Finally since there is a high correlation between neurons,
we filter some more noise through a principal component analysis. From the $|V|$
eigen vectors, we kept $80\%$ of the eigen vectors which have the largest eigen
value. In the context of the challenge, this assumption was deduced from ground
truth of provided networks.

\textcolor{red}{G: Change notations for $f(x^t) = ...$, $g(x^t) = ...$}

@@

If the random variables $X$ follows a joint Gaussian distribution
$p_\mathcal{X} \sim \mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and
covariance matrix $\Sigma$, we have the conditional independence
$X_i \perp X_j | X^{-i,j}$ if the partial correlation coefficient
\[
\rho_{X_i, X_j | X^{-i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}
\]
is equal to zero \citep{koller2009probabilistic}. The partial correlation gives
a connectivity score
$\hat{A}: \hat{a}_{i,j} = \rho_{X_i, X_j | X^{-i,j}}, \forall i, j \in V$ between all
pairs of neuron $i$ and $j$. One could retrieve an estimated graph by thresholding
the matrix $\hat{A}$.
Similarly, we also have $X_i \perp X_j$ if the Pearson correlation
coefficient
\[
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}}
\]
is equal to zero.

@@

% Introduce a custom solution to make the Y matrix asymmetric
% Introduce directivity
\textcolor{red}{Est-ce qu'on ne le ferait pas passer dans l'appendice B? Ou du moins l'équation?}
Since partial correlation is a symmetric measure, the causal mechanism behind the
interaction ($A$ directly causes $B$, $B$ directly causes $A$ or both) can not
be directly inferred\footnote{Also note that two other causal mechanisms might be
implied while having a non-zero partial coefficient: (i) a pair of variables
is induced by a common (hidden) variable; (ii) $A$ (respectively $B$) is
conditionally correlated to a hidden variable affecting $B$ (resp. $A$)
\citep{de2004discovery}.}.
In order to retrieve causal information, we developed an
heuristics based on the following observation. The rise of fluorescence
indicates the activation of a neuron $i$. If a neuron $j$ have also
an increased of fluorescence after a slight delay, this could have been
caused by the neuron $i$. Thus, we count whenever
such events occurred between consecutive time steps
\[
\hat{Z}: z_{ij} = \sum_{t=1}^{T - 1}
    \mathbb{1}(x_j^{t+1} - x_i^t \in \left[f_1, f_2 \right]) -
    \mathbb{1}(x_i^{t+1} - x_j^t \in \left[f_1, f_2 \right]), \forall i, j \in V
\]
where $f_1$ and $f_2$ are parameters of the method.
Finally, the weighted average of correlation matrices is combined with $Z$ such
that $Z$ has a contribution of $0.3\%$ to the prediction of an edge.

@@

% Datasets
To assess our method, we have used four datasets provided by the organizers
of the Connectomics challenge : \textit{normal-1, normal-2, normal-3, normal-4}.
Those datasets are time series of size $T=179500$ of image calcium from simulated
neuronal networks \citep{stetter2012model} with $|V|=1000$ neurons. The networks
to inferred have around 15000 edges and are similar to those
used to rank challenge participants.

% Metrics : ROC AUC and AUPRC
In supervised learning terms, the network inference task can be viewed as a
binary classification task where one has to correctly classify the presence
or the absence of edges of the graph. We assess the accuracy of the
network inference method using the area under the ROC curve (AUROC)
and the area under the precision-recall curve (AUPRC)
\citep{schrynemackers2013protocols}. Since neurons are not self-connected
($(i, i) \not \in E, \forall i \in V$) and some inference methods could
give maximal scores to the diagonal of $\hat{A}$, e.g. Pearson
correlation, we set those elements to the minimum of $\hat{A}$ prior
assessment.

@@

GTE is a method which characterizes a relationship between neurons by an
estimation of the generalized transfer entropy \citep{stetter2012model}. The
goal of this method tries to see if the next time step value is better from
the previous values of this time series or from time series of other neurons.

@@

In previous study \citep{shipley2002cause}, it has been suggested to
consider correlation coefficients of all possible orders,
i.e. conditioned on every subsets of the set of $p-2$ other variables. This
would take into account multiple indirect paths between a given pair of
variables.
